---
title: "README"
output: html_document
---
### Getting and Cleaning Data Project 
#### The instructions below explain all the steps implemented in the script run_analysis, used to generate the tidy data set from the raw data sets

> Running the run_analysis.R script should execute all the steps described bellow smoothly, creating the tiny data set file.

> This script runs with RStudio Version 0.98.1049


**STEP 1** *Merges the training and the test sets to create one data set.*

For  **test data sets** the raw files that we will be working with are:

* subject_test.txt - containing the number that identify each one of 30 the subjects in the study
* X_test.txt - the actual raw data, obtained from the sensor data
* y_test.txt - containing the number used for each activity label as described bellow:
    + 1 = "WALKING"
    + 2 = "WALKING_UPSTAIRS"
    + 3 = "WALKING_DOWNSTAIRS"
    + 4 = "SITTING"
    + 5 = "STANDING"
    + 6 = "LAYING"

In a similar way,
for  **training data sets** the raw files that we will be working with are:

* subject_train.txt - containing the number that identify each one of 30 subjects in the study
* X_train.txt - the actual raw data, obtained from the sensor data
* y_train.txt - containing the number used for each activity label as described above for the test data set.
    
> obs: the sensor data sets will not be used to create our tidy dataset, once both X_test.txt and X_train.txt data sets come from them.

We will work first with the **test dataset** and then we will replicate the same steps to prepare the **train dataset**.

> be sure to have a folder called *project* to contain the UCI HAR Dataset structure on your working directory

The **test files** are located under the UCI HAR Dataset/test folder into the project directory. 
```{r, eval=FALSE}
df_subject_test <- read.table("./project/UCI HAR Dataset/test/subject_test.txt")
df_X_test <- read.table("./project/UCI HAR Dataset/test/X_test.txt")
df_labels_test <- read.table("./project/UCI HAR Dataset/test/y_test.txt")
```
The **train files** are located under the UCI HAR Dataset/train folder into the project directory.
```{r, eval=FALSE}
df_subject_train <- read.table("./project/UCI HAR Dataset/train/subject_train.txt")
df_X_train <- read.table("./project/UCI HAR Dataset/train/X_train.txt")
df_labels_train <- read.table("./project/UCI HAR Dataset/train/y_train.txt")
```

We will work the test data set first and repeat the steps for the train data set later.

We have to merge in a single file the subject ID, activity label and the actual data, considering that all of them are in the proper order. 
Before that, we will change the column names for the activity labels and subject datasets, because they are both named "V1" and this same name is used in the X-test dataset.

```{r, eval=FALSE}
colnames(df_labels_test)[1] <- "activity"
colnames(df_subject_test)[1] <- "subject"
```

Now we can safely merge the 3 datasets...
```{r, eval=FALSE}
df_l_s_X_test <- cbind(df_subject_test,df_labels_test,df_X_test)
```
The dataset df_l_s_X_test now contains all the test data associated with the subject and activity.

Now we repeat the steps above to create the **train dataset**

Changing the column names...
```{r, eval=FALSE}
colnames(df_labels_train)[1] <- "activity"
colnames(df_subject_train)[1] <- "subject"
```
Merging the train datasets...
```{r, eval=FALSE}
df_l_s_X_train <- cbind(df_subject_train,df_labels_train,df_X_train)
```
Now we concatenate verticaly both the **test dataset** and the **train dataset**, for they have the same columns
```{r, eval=FALSE}
df_total <- rbind (df_l_s_X_test,df_l_s_X_train)
```

**STEP 2** *Extracts only the measurements on the mean and standard deviation for each measurement.*

We know that the first column in the df_mean_std dataset is named "subject" , the second is named "activity" and the remaining, with the actual measurement data, goes from "V1" to "V561".
Using the information in the file features.txt, we will select only the variables that are associated with the mean and standard deviation.

Our criteria was to select all the variables that ended with the "mean" or "stdved" sufix.

To extract only the mean and standard deviation for each measurement we will use the dplyr package...
```{r, eval=FALSE}
library(dplyr)
df_mean_std <- select(df_total,subject,activity,V1:V6, V41:V46,V81:V86,V121:V126,V161:V166,V201:V202,
                      V214:V215,V227:V228,V240:V241,V253:V254,V266:V271,V345:V350,
                      V424:V429,V503:V504,V516:V517,V529:V530,V542:V543)
```
Now we have only the measurements needed in the dataset df_mean_std.

**STEP 3** *Uses descriptive activity names to name the activities in the data set*

**IMPORTANT:** We are now cleaning the not longer needed datasets before proceed. This is needed to free memory when dealing with big datasets. Otherwise you can have strange problems occurring in R Studio...
```{r, eval=FALSE}
rm(df_l_s_X_test,df_l_s_X_train,df_X_test, df_X_train, df_total)
```


We then change the numeric values of the activities to their associated labels...
```{r, eval=FALSE}
df_mean_std$activity <- as.character(df_mean_std$activity) # transforming activity from integer type to character type
df_mean_std$activity[df_mean_std$activity == 1] <- "WALKING"
df_mean_std$activity[df_mean_std$activity == 2] <- "WALKING_UPSTAIRS"
df_mean_std$activity[df_mean_std$activity == 3] <- "WALKING_DOWNSTAIRS"
df_mean_std$activity[df_mean_std$activity == 4] <- "SITTING"
df_mean_std$activity[df_mean_std$activity == 5] <- "STANDING"
df_mean_std$activity[df_mean_std$activity == 6] <- "LAYING"
```

**STEP 4** *From the data set in step 3, creates a second, independent tidy data set with the average of each variable for each activity and each subject.*

In this step, we will group the data for activity and subject, obtaining the mean by using dplyer package again and write the tidy file.

```{r, eval=FALSE}
df_grouped <- group_by(df_mean_std, activity, subject)
df_tidy <- summarise(df_grouped,mean(V1), mean(V2),mean(V3), mean(V4),mean(V5), mean(V6),
                 mean(V41), mean(V42),mean(V43), mean(V44),mean(V45), mean(V46),
                 mean(V81), mean(V82),mean(V83), mean(V84),mean(V85), mean(V86),
                 mean(V121), mean(V122),mean(V123), mean(V124),mean(V125), mean(V126),
                 mean(V161), mean(V162),mean(V163), mean(V164),mean(V165), mean(V166),
                 mean(V201), mean(V202),
                 mean(V214), mean(V215),
                 mean(V227), mean(V228),
                 mean(V240), mean(V241),
                 mean(V253), mean(V254),
                 mean(V266), mean(V267),mean(V268), mean(V269),mean(V270), mean(V271),
                 mean(V345), mean(V346),mean(V347), mean(V348),mean(V349), mean(V350),
                 mean(V424), mean(V425),mean(V426), mean(V427),mean(V428), mean(V429),
                 mean(V503), mean(V504),
                 mean(V516), mean(V517),
                 mean(V529), mean(V530),
                 mean(V542), mean(V543)
                 )
```

**STEP 4** *Appropriately labels the data set with descriptive variable names.* 

Now, in this final step, using again the features.txt file as a guide, we will change the current column names, to more descriptive names.

I am pretty sure that with more available time I should come up with a loop or some package to do this easier,
but, anyway, the first version is intended to get the job done, right?

```{r, eval=FALSE}
# creating descriptive names for variables in the range V1:V6
names(df_mean_std)[names(df_mean_std) == 'V1'] <- "BodyAcceleration_MeanTimeDirX"
names(df_mean_std)[names(df_mean_std) == 'V2'] <- "BodyAcceleration_MeanTimeDirY"
names(df_mean_std)[names(df_mean_std) == 'V3'] <- "BodyAcceleration_MeanTimeDirZ"
names(df_mean_std)[names(df_mean_std) == 'V4'] <- "BodyAcceleration_StddevTimeDirX"
names(df_mean_std)[names(df_mean_std) == 'V5'] <- "BodyAcceleration_StddevTimeDirY"
names(df_mean_std)[names(df_mean_std) == 'V6'] <- "BodyAcceleration_StddevTimeDirZ"
# creating descriptive names for variables in the range V41:V46
names(df_mean_std)[names(df_mean_std) == 'V41'] <- "GravityAcceleration_MeanTimeDirX"
names(df_mean_std)[names(df_mean_std) == 'V42'] <- "GravityAcceleration_MeanTimeDirY"
names(df_mean_std)[names(df_mean_std) == 'V43'] <- "GravityAcceleration_MeanTimeDirZ"
names(df_mean_std)[names(df_mean_std) == 'V44'] <- "GravityAcceleration_StddevTimeDirX"
names(df_mean_std)[names(df_mean_std) == 'V45'] <- "GravityAcceleration_StddevTimeDirY"
names(df_mean_std)[names(df_mean_std) == 'V46'] <- "GravityAcceleration_StddevTimeDirZ"
# creating descriptive names for variables in the range V81:V86
names(df_mean_std)[names(df_mean_std) == 'V81'] <- "BodyAccelerationJerk_MeanTimeDirX"
names(df_mean_std)[names(df_mean_std) == 'V82'] <- "BodyAccelerationJerk_MeanTimeDirY"
names(df_mean_std)[names(df_mean_std) == 'V83'] <- "BodyAccelerationJerk_MeanTimeDirZ"
names(df_mean_std)[names(df_mean_std) == 'V84'] <- "BodyAccelerationJerk_StddevTimeDirX"
names(df_mean_std)[names(df_mean_std) == 'V85'] <- "BodyAccelerationJerk_StddevTimeDirY"
names(df_mean_std)[names(df_mean_std) == 'V86'] <- "BodyAccelerationJerk_StddevTimeDirZ"
# creating descriptive names for variables in the range V121:V126
names(df_mean_std)[names(df_mean_std) == 'V121'] <- "BodyGyroscopic_MeanTimeDirX"
names(df_mean_std)[names(df_mean_std) == 'V122'] <- "BodyGyroscopic_MeanTimeDirY"
names(df_mean_std)[names(df_mean_std) == 'V123'] <- "BodyGyroscopic_MeanTimeDirZ"
names(df_mean_std)[names(df_mean_std) == 'V124'] <- "BodyGyroscopic_StddevTimeDirX"
names(df_mean_std)[names(df_mean_std) == 'V125'] <- "BodyGyroscopic_StddevTimeDirY"
names(df_mean_std)[names(df_mean_std) == 'V126'] <- "BodyGyroscopic_StddevTimeDirZ"
# creating descriptive names for variables in the range V161:V166
names(df_mean_std)[names(df_mean_std) == 'V161'] <- "BodyGyroscopicJerk_MeanTimeDirX"
names(df_mean_std)[names(df_mean_std) == 'V162'] <- "BodyGyroscopicJerk_MeanTimeDirY"
names(df_mean_std)[names(df_mean_std) == 'V163'] <- "BodyGyroscopicJerk_MeanTimeDirZ"
names(df_mean_std)[names(df_mean_std) == 'V164'] <- "BodyGyroscopicJerk_StddevTimeDirX"
names(df_mean_std)[names(df_mean_std) == 'V165'] <- "BodyGyroscopicJerk_StddevTimeDirY"
names(df_mean_std)[names(df_mean_std) == 'V166'] <- "BodyGyroscopicJerk_StddevTimeDirZ"
# creating descriptive names for variables in the range V201:V202
names(df_mean_std)[names(df_mean_std) == 'V201'] <- "BodyAcceleration_MeanTimeMagnitude"
names(df_mean_std)[names(df_mean_std) == 'V202'] <- "BodyAcceleration_StddevTimeMagnitude"
# creating descriptive names for variables in the range V214:V215
names(df_mean_std)[names(df_mean_std) == 'V214'] <- "GravityAcceleration_MeanTimeMagnitude"
names(df_mean_std)[names(df_mean_std) == 'V215'] <- "GravityAcceleration_StddevTimeMagnitude"
# creating descriptive names for variables in the range V227:V228
names(df_mean_std)[names(df_mean_std) == 'V227'] <- "BodyAccelerationJerk_MeanTimeMagnitude"
names(df_mean_std)[names(df_mean_std) == 'V228'] <- "BodyAccelerationJerk_StddevTimeMagnitude"
# creating descriptive names for variables in the range V240:V241
names(df_mean_std)[names(df_mean_std) == 'V240'] <- "BodyGyroscopic_MeanTimeMagnitudee"
names(df_mean_std)[names(df_mean_std) == 'V241'] <- "BodyGyroscopic_StddevTimeMagnitude"
# creating descriptive names for variables in the range V253:V254
names(df_mean_std)[names(df_mean_std) == 'V253'] <- "BodyGyroscopicJerk_MeanTimeMagnitude"
names(df_mean_std)[names(df_mean_std) == 'V254'] <- "BodyGyroscopicJerk_StddevTimeMagnitude"
# creating descriptive names for variables in the range V266:V271
names(df_mean_std)[names(df_mean_std) == 'V266'] <- "BodyAcceleration_MeanFfourierDirX"
names(df_mean_std)[names(df_mean_std) == 'V267'] <- "BodyAcceleration_MeanFfourierDirY"
names(df_mean_std)[names(df_mean_std) == 'V268'] <- "BodyAcceleration_MeanFfourierDirZ"
names(df_mean_std)[names(df_mean_std) == 'V269'] <- "BodyAcceleration_StddevFfourierDirX"
names(df_mean_std)[names(df_mean_std) == 'V270'] <- "BodyAcceleration_StddevFfourierDirY"
names(df_mean_std)[names(df_mean_std) == 'V271'] <- "BodyAcceleration_StddevFfourierDirZ"
# creating descriptive names for variables in the range V345:V350
names(df_mean_std)[names(df_mean_std) == 'V345'] <- "BodyAccelerationJerk_MeanFfourierDirX"
names(df_mean_std)[names(df_mean_std) == 'V346'] <- "BodyAccelerationJerk_MeanFfourierDirY"
names(df_mean_std)[names(df_mean_std) == 'V347'] <- "BodyAccelerationJerk_MeanFfourierDirZ"
names(df_mean_std)[names(df_mean_std) == 'V348'] <- "BodyAccelerationJerk_StddevFfourierDirX"
names(df_mean_std)[names(df_mean_std) == 'V349'] <- "BodyAccelerationJerk_StddevFfourierDirY"
names(df_mean_std)[names(df_mean_std) == 'V350'] <- "BodyAccelerationJerk_StddevFfourierDirZ"
# creating descriptive names for variables in the range V424:V429
names(df_mean_std)[names(df_mean_std) == 'V424'] <- "BodyGyroscopic_MeanFfourierDirX"
names(df_mean_std)[names(df_mean_std) == 'V425'] <- "BodyGyroscopic_MeanFfourierDirY"
names(df_mean_std)[names(df_mean_std) == 'V426'] <- "BodyGyroscopic_MeanFfourierDirZ"
names(df_mean_std)[names(df_mean_std) == 'V427'] <- "BodyGyroscopic_StddevFfourierDirX"
names(df_mean_std)[names(df_mean_std) == 'V428'] <- "BodyGyroscopic_StddevFfourierDirY"
names(df_mean_std)[names(df_mean_std) == 'V429'] <- "BodyGyroscopic_StddevFfourierDirZ"
# creating descriptive names for variables in the range V503:V504
names(df_mean_std)[names(df_mean_std) == 'V503'] <- "BodyAcceleration_MeanFfourierMagnitude"
names(df_mean_std)[names(df_mean_std) == 'V504'] <- "BodyAcceleration_StddevFfourierMagnitude"
# creating descriptive names for variables in the range V516:V517
names(df_mean_std)[names(df_mean_std) == 'V516'] <- "BodyAccelerationJerk_MeanFfourierMagnitude"
names(df_mean_std)[names(df_mean_std) == 'V517'] <- "BodyAccelerationJerk_StddevFfourierMagnitude"
# creating descriptive names for variables in the range V529:V530
names(df_mean_std)[names(df_mean_std) == 'V529'] <- "BodyGyroscopic_MeanFfourierMagnitude"
names(df_mean_std)[names(df_mean_std) == 'V530'] <- "BodyGyroscopic_StddevFfourierMagnitude"
# creating descriptive names for variables in the range V542:V543
names(df_mean_std)[names(df_mean_std) == 'V542'] <- "BodyGyroscopicJerk_MeanFfourierMagnitude"
names(df_mean_std)[names(df_mean_std) == 'V543'] <- "BodyGyroscopicJerk_StddevFfourierMagnitude"
```

Finaly we can save our data set to a text file...
```{r, eval=FALSE}
write.table(df_tidy, file = "./data/tidy_project.txt",sep = ",",row.name = FALSE)
```

> the file can be easily read if imported into an Excel spreadsheet.

If you want to check the if the values generated are correct, try the code below (that I keep commented in the script file),
selecting some random values and comparing them with the values into the tidy_project.txt file.
If you import the file into a Excel spreadsheet it will be more confortable to work.
```{r, eval=FALSE}
# uncomment the code below to test the final file generated. 
# remember that the variable names are not changed in this dataset (V1, V2, etc).
# but as it is in the same order, it is easy to select some values and check against the data file.
# you can import the text file to Excel to make the things easier
x <- df_mean_std$V2[df_mean_std$activity == "WALKING" & df_mean_std$subject == 14]
mean(x)
```
Thanks!!

